---
title: "Multistate capture-recapture and mark-recovery models"
author: "Frederic Barraquand"
date: "February 14, 2019"
output:
  pdf_document: default
  number_sections: yes
  word_document: null
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
csl: ecology.csl
bibliography: /home/frederic/Dropbox/database/total.bib
---

## Intro

We need a model combining both CMR and mark-recovery (dead individuals). This versions starts in 1981 because we have the covariates starting that year. 

## Create a matrix of capture histories

```{r}
dringed<-read.csv("data/Gyrs_ringed.csv")
dringed20<-read.csv("data/RingingDataAdditionJan2020.csv")
dringed<-rbind(dringed,dringed20)
head(dringed) # What's in the data table
length(unique(dringed$Ring_No)) #How many unique bird IDs - 1816 (vs 1653 previously)

### Removing below 1981
nrow(dringed[dringed$Ringed_year<1981,]) #25 entries, that's a pity. 
birds_ID_before81 = dringed$Ring_No[dringed$Ringed_year<1981]
dringed=dringed[!(dringed$Ringed_year<1981),]
#check they disappeared. 
dringed[dringed$Ringed_year<1981,]
#OK
nrow(dringed)
```

OK, so we have 1791 unique IDs (with the removal). Let's compare to how many gyrs have been recovered. 

```{r}
drecov<-read.csv("data/Gyrs_recovered.csv")
drecov20<-read.csv("data/EncounterDataAdditionJan2020.csv")
drecov<-rbind(drecov,drecov20)
head(drecov)
length(unique(drecov$Ring_No)) #How many unique bird IDs in Gyrs_recovered.csv (293 vs 270 previously)

## removing now the birds before 1981
drecov[drecov$Ring_No %in%birds_ID_before81,] #7 recoveries
drecov=drecov[!(drecov$Ring_No %in%birds_ID_before81 ),]
drecov[drecov$Ring_No %in%birds_ID_before81,] #sucessfully removed. 
sort(unique(drecov$Year_recorded))
sort(unique(dringed$Ringed_year)) #all good. 
```

Actually we need a mixture of mark-recovery and CMR. p. 295 Kery and Schaub. 
In our case, not all individuals in the recently dead state move to the "dead" state at t+1. This is because recently dead individuals can be recovered still after t+1 (although we should probably set a prior that makes this probability fairly high).  

Let's compute capture histories (without distinguish young and old for now?)

```{r}
Birds_ringed = unique(dringed$Ring_No)
Birds_recov = unique(drecov$Ring_No)
min_year = min(dringed$Ringed_year)
max_year = max(dringed$Ringed_year)
min_year
max_year
### check
min(drecov$Year_recorded)
max_year = max(drecov$Year_recorded) #because it's 2019
timespan = min_year:max_year
time_index = 1:length(timespan)
time_index_max = length(timespan)
CH = matrix(NA,nrow=nrow(dringed),ncol=time_index_max)

dringed$First_recov_date = NA
dringed$First_recov_year = NA
dringed$First_recov_type = NA

for (i in 1:nrow(dringed))
{
    ## if it's ringed, we put a 1 on capture year
    year_ringed = dringed$Ringed_year[i]-min_year+1
    CH[i,year_ringed] = 1
  if (dringed$Ring_No[i] %in% Birds_recov)
  {
    ## Keep in mind they can be recorded several times
    year_recov = drecov$Year_recorded[drecov$Ring_No == as.character(dringed$Ring_No[i])]
    dates_recov = drecov$Date_recovered[drecov$Ring_No == as.character(dringed$Ring_No[i])]
    type_recov = drecov$Condition_Euring_code[drecov$Ring_No == as.character(dringed$Ring_No[i])]
    
    dringed[i,]$First_recov_year = year_recov[1]
    dringed[i,]$First_recov_date = as.character(as.Date(dates_recov[1],format="%m/%d/%Y"))
    dringed[i,]$First_recov_type = type_recov[1]
    
    for (d in 1:length(year_recov)){
        year_recov_focal = year_recov[d]-min_year+1

         if  (type_recov[d] %in% c(4,7,8)){ #live recovery
           CH[i,year_recov_focal] = 1
         } 
         else {
           CH[i,year_recov_focal] = 2
           
           }
      }#end of for loop on all dates where the bird was recorded

  }#end of if condition on whether the bird was captured
  else {
    if (year_ringed<time_index_max)
        {
          CH[i,(year_ringed+1):time_index_max] = 3
        }
    #else nothing, since we do not have to add what happens afterwards
  }
  
}
CH[is.na(CH)]=3

### Plotting the capture histories
library('RColorBrewer')
library('fields')
cols = brewer.pal(3,"RdBu")
rf <- colorRampPalette(cols)   # make colors
# voir aussi  http://www.sthda.com/french/wiki/couleurs-dans-r
image.plot(1:nrow(dringed),timespan,CH,col=cols)

#Verif of data by blocks
time1=1
time2=250
image.plot(time1:time2,timespan,CH[time1:time2,],col=cols)

time1=250
time2=500
image.plot(time1:time2,timespan,CH[time1:time2,],col=cols)

time1=500
time2=750
image.plot(time1:time2,timespan,CH[time1:time2,],col=cols)
## A number of individuals are either recaptured dead the same year or there is a date problem



time1=750
time2=1000
image.plot(time1:time2,timespan,CH[time1:time2,],col=cols)

time1=1000
time2=1250
image.plot(time1:time2,timespan,CH[time1:time2,],col=cols)

time1=1250
time2=nrow(dringed)
image.plot(time1:time2,timespan,CH[time1:time2,],col=cols)

```

Troubleshoot cases where the year of marking is also the year where the bird has been recorded dead

```{r}
### Create data structure where the recovery information is appended to ringed birds dataframe
dringed_recovered = subset(dringed,(dringed$Ring_No %in% Birds_recov) & (dringed$Ringed_year==dringed$First_recov_year))
dringed_pb = subset(dringed_recovered,dringed_recovered$First_recov_type==2)
### Actually it's *more* of a Pb if the recovery type is 2 but it is also a problem if it is 
dringed_pb = dringed_recovered
head(dringed_pb)
nrow(dringed_pb) #33 before -- 112 now!
dringed_pb
write.csv(dringed_pb,file="RingedAndRecovered_sameYear_jan2020.csv")

### Are all the birds ringed in June-July? 
table(dringed$Ringed_month)
### Answer is May-June-July. We can say that they have been observed the next year to be dead if they are dead in the year of the ringing, but not if they are alive. 
### It means that if they have been seen alive at 6 month, we don't know the status the next year. But on the other hand, only death is permanent. 
### Sounds reasonable to write a 2 the next year in that case. 
### Let's do that

CH = matrix(NA,nrow=nrow(dringed),ncol=time_index_max)

for (i in 1:nrow(dringed))
{
    ## if it's ringed, we put a 1 on capture year
    year_ringed = dringed$Ringed_year[i]-min_year+1
    CH[i,year_ringed] = 1
  if (dringed$Ring_No[i] %in% Birds_recov)
  {
    ## Keep in mind they can be recorded several times
    year_recov = drecov$Year_recorded[drecov$Ring_No == as.character(dringed$Ring_No[i])]
    dates_recov = drecov$Date_recovered[drecov$Ring_No == as.character(dringed$Ring_No[i])]
    type_recov = drecov$Condition_Euring_code[drecov$Ring_No == as.character(dringed$Ring_No[i])]
    
    dringed[i,]$First_recov_year = year_recov[1]
    dringed[i,]$First_recov_date = as.character(as.Date(dates_recov[1],format="%m/%d/%Y"))
    dringed[i,]$First_recov_type = type_recov[1]
    
    for (d in 1:length(year_recov)){
        year_recov_focal = year_recov[d]-min_year+1

         if  (type_recov[d] %in% c(4,7,8)){ #live recovery
           CH[i,year_recov_focal] = 1
         } 
         else {
           # Asks whether the year of the ring is the year of recovery
           if(dringed[i,]$Ringed_year==dringed[i,]$First_recov_year){#Yes
             #Then keep the 1 and put 2 the next year, since the individual died through that year
             CH[i,year_recov_focal] = 1
             if ((year_recov_focal+1)<time_index_max){CH[i,year_recov_focal+1] = 2}
           }
           else { #the ring of the year is not that of the recovery
             CH[i,year_recov_focal] = 2
           }
        
           }
      }#end of for loop on all dates where the bird was recorded

  }#end of if condition on whether the bird was captured
  else {
    #birds that are never captured again
    if (year_ringed<time_index_max)
        {
          CH[i,(year_ringed+1):time_index_max] = 3
        }
    #else nothing, since we do not have to add what happens afterwards
  }
  
}

CH[is.na(CH)]=3

```

Verification of the data by blocks of individuals

```{r}

#Verif of data by blocks
time1=1
time2=250
image.plot(time1:time2,timespan,CH[time1:time2,],col=cols)

time1=250
time2=500
image.plot(time1:time2,timespan,CH[time1:time2,],col=cols)

time1=500
time2=750
image.plot(time1:time2,timespan,CH[time1:time2,],col=cols)

time1=750
time2=1000
image.plot(time1:time2,timespan,CH[time1:time2,],col=cols)

time1=1000
time2=1250
image.plot(time1:time2,timespan,CH[time1:time2,],col=cols)

time1=1250
time2=nrow(dringed)
image.plot(time1:time2,timespan,CH[time1:time2,],col=cols)

### All good. 
write.csv(CH,file = "data/CaptureHistories1981til2020.csv")

```

Now we need to write down the status of the individuals, what they are at the first stage. 

```{r}
table(dringed$EurAgeCode)
table(dringed$EurAgeText)
# So we output a vector 
dringed$EurAgeCode[dringed$EurAgeCode==8]=2
stagemarking = dringed$EurAgeCode
write.csv(stagemarking,file = "data/stagemarking1981til2020.csv")

#We also need the year of first marking
fmarking = dringed$First_recov_year
write.csv(fmarking ,file = "data/fmarking1981til2020.csv")

```

