---
title: "Preliminary analysis of the Icelandic Gyrfalcon CMR dataset - v2"
author: "Frederic Barraquand, Olafur Nielsen"
date: "February 22, 201_"
output:
  pdf_document: default
  number_sections: yes
  word_document: null
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
csl: ecology.csl
bibliography: /home/frederic/Dropbox/database/total.bib
---

Here I attempt to explore the CMR dataset, before we go to inferential models to estimate survival and perhaps link this to an integrated pop. model. 

## Basic statistics

How many birds have been ringed? 

```{r}
dringed<-read.csv("data/Gyrs_ringed.csv")
head(dringed) # What's in the data table
length(unique(dringed$Ring_No)) #How many unique bird IDs
```

OK, so we have 1653 unique IDs. Let's compare to how many gyrs have been recovered. 

```{r}
drecov<-read.csv("data/Gyrs_recovered.csv")
head(drecov)
length(unique(drecov$Ring_No)) #How many unique bird IDs in Gyrs_recovered.csv
```

Let's now analyse the number of birds that have been ringed as a function of the year 

```{r,echo=FALSE} 
#dringed$Year=format(as.Date(dringed$Ringed_date, format="%d/%m/%Y"),"%Y") seems to create NAs
dringed$Year=format(as.Date(dringed$Ringed_date, format="%m/%d/%Y"),"%Y") ### Month was first in the data
years_ringed=sort(unique(as.numeric(na.omit(dringed$Year))))
ringed_birds_per_year=years_ringed # initializing
for (k in 1:length(years_ringed)){
  ringed_birds_per_year[k]=length(unique(dringed$Ring_No[dringed$Year==years_ringed[k]]))
}
plot(years_ringed,ringed_birds_per_year,type="h",lwd=5,col="red")
#ggplot(data=,aes(x=years_ringed, y=ringed_birds_per_year)) + geom_bar(stat="identity")
```

We now analyse the patterns of recovery (and resighting)

```{r,echo=FALSE} 
drecov$Year=format(as.Date(drecov$Date_recovered, format="%m/%d/%Y"),"%Y")
years_of_recovery=sort(unique(as.numeric(na.omit(drecov$Year))))
birds_recovered_per_year=years_of_recovery# initializing
for (k in 1:length(years_of_recovery)){
  birds_recovered_per_year[k]=length(unique(drecov$Ring_No[drecov$Year==years_of_recovery[k]]))
}
plot(years_of_recovery,birds_recovered_per_year,type="h",lwd=5,col="blue")
```

Now how many birds have been seen several times? We see below that very few birds have been recovered more than once (also this tends to be recent?). 

----
**Question: when did the change from recovery of dead birds to resighting of live birds occur?**
----
I can infer the earliest date at which a bird has been resighted a second time, but given few birds have been resighted, but it might be preferable to allow later for a change in protocol (in the CMR models) using a predefined time (we can use several if unsure). I have noted 2006 earlier but I am unsure. 

$~$

```{r,echo=FALSE} 
bird_recovered_list=unique(drecov$Ring_No)
ntimes_recovered=rep(0,length(bird_recovered_list))
for (k in 1:length(bird_recovered_list)){
  ntimes_recovered[k]=sum(drecov$Ring_No==bird_recovered_list[k])
}
plot(bird_recovered_list,ntimes_recovered,type="o",lwd=1,col="black",xlab="Bird ID",ylab= "Ntimes recovered")
table(ntimes_recovered)
```

This is coherent with the info given by Oli -- 1 bird recovered 4 times, 3 birds 3 times, 11 birds two times and the rest 1 time. 

Thus for all practical purposes, we can consider that such data consists mainly of individual that are either recovered or not recovered/resighted. We'll now turn to whether the individuals have been found live or dead, and how this varies in time. 

Also whether the recoveries of dead birds are of young vs. adult birds (which may provide quick and dirty estimates of survival rates, at least for the youngs...)

```{r} 
## these are all the recovery codes that we have (top row), and how many of those we have in the dataset (bottom row).

table(drecov$Condition_Euring_code) 

### [previous comment: For instance, we can see 15 read colourmarks. 
### We need to simplify this complexity to some extent by aggregating some values. ]

```

 *Now we need some more info on the codes*
 
 These are defined according to The EURING EXCHANGE Code 2000+ 
https://euring.org/files/documents/E2000PLUSExchangeCodeV117.pdf

| Condition Code | Meaning                                            | 
|--------------- |:--------------------------------------------------:|
| 0              | Condition completely unknown                       |
| 1              | Dead but no information on how recently died/killed|
| 2              | Freshly dead – within about a week                 |
| 3              | dead for > a week. If >>, use 9 in Accuracy of Date, and 3 here|
| 4              | Found sick/wounded and released afterwards  |
| 5              | Found sick/wounded and NOT released afterwards  |
| 6              | Alive and probably healthy but taken into captivity. |
| 7              | Alive and probably healthy and certainly released (ring seen without the bird having being caught). |
| 8              | Alive and probably healthy and released by a ringer |
| 9              | Alive and probably healthy but ultimate fate of bird is not known|

Thus categories 4, 7 and 8 (respectively  8, 31 and 21 birds) correspond to cases where the bird is 'released' alive, i.e. effectively resighted rather than recovered (when dead). 


```{r,echo=TRUE}
drecov$Date_recovered[drecov$Condition_Euring_code %in% c(4,7,8)]

drecov$Year[drecov$Condition_Euring_code %in% c(4,7,8)]

vec_year_recovered_live = as.numeric(drecov$Year[drecov$Condition_Euring_code %in% c(4,7,8)])
hist(vec_year_recovered_live,breaks=20,xlab = "Years of live recovery", main = "Live recaptures (Euring 4,7,8)")

drecov$Date_recovered[drecov$Condition_Euring_code %in% c(7,8)]

drecov$Date_recovered[drecov$Condition_Euring_code ==7]

drecov$Date_recovered[drecov$Condition_Euring_code ==8]

### Check not a problem of date recording
drecov$Year[(drecov$Condition_Euring_code %in% c(4,7,8))&(drecov$Accurancy_of_date_Euring_code!=9)]

drecov$Accurancy_of_date_Euring_code[drecov$Condition_Euring_code %in% c(4,7,8)]


```


```{r}
table(drecov$Accurancy_of_date_Euring_code) 

## Check the birds whose dying dates are poorly known 
drecov$Year[(drecov$Accurancy_of_date_Euring_code %in% c(8,9))]
drecov$Condition_Euring_code[(drecov$Accurancy_of_date_Euring_code %in% c(8,9))]

## New dataframe for looking into those patterns
long_dead_birds=drecov[(drecov$Accurancy_of_date_Euring_code %in% c(8,9)),]
head(long_dead_birds)
### Possible to correct the date using the When.dead column
nrow(long_dead_birds)
long_dead_birds$When.dead
## For which do we have information?
long_dead_birds$When.dead[grepl("D",long_dead_birds$When.dead)]
long_dead_birds$Ring_No[grepl("D",long_dead_birds$When.dead)]

```

| Accuracy of date Code | Meaning                                            | 
|---------------------- |:--------------------------------------------------:|
| 0                     | Accurate to the day       |
| 1                     | Accurate to within 1 day either side of date coded. |
| 2                     | Accurate to within 3 days either side of date coded. |
| 3                     | Accurate to within 1 week either side of date coded. |
| 4                     | Accurate to within 2 weeks either side of date coded. |
| 5                     | Accurate to within 6 weeks either side of date coded. |
| 6                     | Accurate to within 3 months either side of date coded. |
| 7                     | Accurate to within 6 months either side of date coded. |
| 8                     | Accurate to within some years only |
| 9                     | Date of earliest possible use of ring (EURING for details) |

 
 For the circumstances, see the EURING pdf. Here are the numbers for the various categories
 
```{r}
table(drecov$Circumstances_Euring_code) 
```

 
 -------------------------------- Old codes --------------------------------------


| AGE Code      | Meaning                     | 
|-------------- |:---------------------------:|
| 100           | unfledged young             |
| 101           | unfledged young at the nest |
| 501           | adult at the nest           |  


 
| RECOVERY Code    | Meaning                          | 
|------------------|:--------------------------------:|
| 100              | found dead                       |
| 120              | found dead for a long time       |
| 121              | found dead with one tag only?    |
| 146              | found dead just outside the nest |
| 5700             | read colourmark                  |
| 981,996          | found injured and had to kill it |




